Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013

### 2017-07-04 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd

---
title: Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

```{r load_data, echo=F}
BlackWhite_results <- read.csv("../Results2/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map2(BlackWhite_results, "state", "stabbrs")
cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))

age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
age_cod_results_infants_AOC$stabbrs <- state.abb[match(age_cod_results_infants_AOC$state, state.name)]
age_cod_results_infants_AOC$stabbrs[age_cod_results_infants_AOC$state == "Washington DC"] <- "DC"
age_cod_results_infants_AOC <- reorder.as.map2(age_cod_results_infants_AOC, "state", "stabbrs")

@@

```{r summarize_results_for_abstract, echo = F}
cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% select(fixed.from.RE) 
cvd.female.2013 <- format(round(cvd.female.2013, 1), nsmall = 1)

ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)

can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(max = round(max(fixed.from.RE), 1))
can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))

inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% select(fixed.from.RE) %>% round(1)
```

Objective: To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across states over the last half-century, and identify divergent states.

@@

Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r cvd.male.1969` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r formatC(as.numeric(cvd.female.2013), digits = 1)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in 1988) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

Conclusions: Life expectancy for both races has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.

@@

- This study provides the foundation for future work in which the estimated trends for each state can be used to study the effects of state-level policy.

OR

@@

- Due to data supression when death counts were small, we did not consider more specific causes of death such as homocide and drug overdose.

@@

###Results
```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  labs(fill = "Census Division") +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  labs(fill = "Census Division") +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```

```{r plot_both, eval=F, echo=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ stabbrs.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

@@

**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state. The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines. These lines are overlaid with the meta-estimated national trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S3-S14**).

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 1A**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 1B**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

@@

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**). Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**). For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 2B**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

@@

Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**). Recently, there was a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**). As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 3A and 3B**). As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress. Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap.

@@

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**). Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**). This spike was greater among men, especially blacks, and among states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

@@

```{r, echo = F}
male.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Male", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

female.endpoints.aoc <- cod_decomp_results %>% filter(year %in% c(1969, 2013), sex =="Female", COD == "All other causes", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

male.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )

female.infants.endpoints.aoc <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% transmute(round(fitted.RE - RE.estimates, 2) )
```

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**). The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally. We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**). We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969. In 2013, this proportion reached 53%. Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.

@@

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the mid-West, which had been the receiving point of “the Great Migration” for improved economic opportunities in the mid-20th Century.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^16^

@@

###Appendix
**Table S1: Cause of death groupings**

| Broad cause of death     | Included causes of death                                                     |
|--------------------------|------------------------------------------------------------------------------|
| Cancers                   | Lip; Tongue; Salivary gland; Floor of mouth; Gum and other mouth; Nasopharnyx; Tonsil; Oropharynx; Hypopharynx; Other oral cavity and pharynx; Esophagus; Stomach; Small intestine; Colon excluding rectum; Rectum and rectosigmoid junction; Anus, anal canal and anorectum; Liver; Intrahepatic bile duct; Gallbladder; Other biliary; Pancreas; Retroperitoneum; Peritoneum, Omentum and mesentery; Other digestive organs; Nose, nasal cavity and middle ear; Larynx; Lung and bronchus; Pleura; Trachea, Mediastinum and other respiratory organs; Bones and joints; Soft tissue including heart; Melanoma of the skin; Non-melanoma skin; Breast; Cervix uteri; Corpus uteri; Uterus, NOS; Ovary; Vagina; Vulva; Other female genital organs; Prostate; Testis; Penis; Other male genital organs; Urinary bladder; Kidney and renal pelvis; Ureter; Other urinary organs; Eye and orbit; Brain and other nervous system; Thyroid; Other endocrine including thymus; Hodkin lymphoma; Myeloma; Acute lymphocytic leukemia; Chronic lymphocytic leukemia; Other lymphocytic leukemia; Acute myeloid leukemia; Acute monocytic leukemia; Chronic myeloid leukemia; Other myeliod/monocytic leukemia; Other acute leukemia; Aleukemic, Subleukemic and NOS; Miscellaneous malignant cancer. |
| Cardiovascular disease   | Diseases of heart; Hypertension without heart disease; Cerebrovascular diseases; Atherosclerosis; Aortic aneurysm and dissection; Other diseases of arteries, arterioles, capillaries. | 
| Communicable disease     | Tuberculosis; Syphilis; HIV (1987+), Septicemia; Other infectious and parasitic diseases; Pneumonia and influenza |
| Non-communicable disease | Diabetes Mellitus; Alzheimers (ICD9 and 10 only); Chronic obstructive pulmonary disease and allied conditions; Stomach and duodenal ulcers; Chronic liver disease and cirrhosis; Nephritis, Nephrotic Syndrome and Nephrosis.  |
| Injuries                 | Accidents and Adverse Effects; Suicide and Self-inflicted Injury; Homicide and Legal Intervention. |
| All other causes         | Complications of pregnancy, childbirth, puerperium; Congenital anomalies; Certain conditions originating in perinatal period; Symptoms, signs, and ill-defined conditions; Other cause of death. |

**Stacked charts - not to be included in the submission**

**Men**

```{r stacked_trends_male, echo=F, fig.height=9, fig.width=11}
plot1 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male"), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot1$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot1$x$data[[i]]$line$width <- 1
  }
}

plot1

plot2 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Male" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
  geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Male" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

for(i in 1:400){
  if(plot2$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot2$x$data[[i]]$line$width <- 1
  }
}

plot2
```

**Women**

```{r stacked_trends_female, echo=F, fig.height=9, fig.width=11}
plot3 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female"), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female"), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (1969-2013)"))

for(i in 1:400){
  if(plot3$x$data[[i]]$line$color == "rgba(0,0,0,1)"){
    plot3$x$data[[i]]$line$width <- 1
  }
}

plot3

plot4 <- ggplotly(ggplot(subset(cod_decomp_results, sex == "Female" & year > 2009), aes(x = year, y = COD_cont_yrs_mean)) + 
                      geom_area(aes(fill = COD_rev), stat = "identity") +
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 2009)) + geom_hline(aes(yintercept = -0.4)) + #the axes
  geom_line(data = subset(BlackWhite_results, sex == "Female" & year > 2009), aes(x = year, y = LE_wbgap_mean)) +
  facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) + 
  theme_classic(base_size = 14) + 
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Contribution to the life expectancy gap (%)") +
  xlab("Year (2009-2013)"))

>>>>

---
title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

```{r load_data, echo=F}
cod_decomp_results <- read.csv("../Results2/cod_decomp_results_cubic.csv")
cod_decomp_results <- reorder.as.map2(cod_decomp_results, "state", "stabbrs")
cod_decomp_results$COD_rev <- factor(cod_decomp_results$COD, levels = c("Cardiovascular", "Cancers", "All other causes",
                                                                        "Communicable", "Non-communicable",  "Injuries"))

age_cod_results_infants_AOC <- read.csv("../Results2/age_cod_results_infants_AOC_cubic.csv")
```

@@

```{r summarize_results_for_abstract, echo = F}
cvd.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>%
  ungroup() %>% select(fixed.from.RE) %>% round(1)
cvd.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
cvd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
cvd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Cardiovascular" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

ncd.male.average <- cod_decomp_results %>% filter(sex == "Male" & COD == "Non-communicable" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))
ncd.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>%  select(fixed.from.RE) %>% round(1)
ncd.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Non-communicable" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

can.male.max <- cod_decomp_results %>% filter(sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))
can.female.average <- cod_decomp_results %>% filter(sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  select(fixed.from.RE) %>% summarize(mean = round(mean(fixed.from.RE), 1))

inj.male.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
inj.male.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Male" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
inj.female.1969 <- cod_decomp_results %>% filter(year == 1969 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
inj.female.2013 <- cod_decomp_results %>% filter(year == 2013 & sex == "Female" & COD == "Injuries" & state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)
```

Objective: To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.

@@

Results: Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

Conclusions: Life expectancy for blacks and whites has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.

@@

@@

- The outcome of life expectancy integrates mortality rates across the entire age distribution making it sensitive to changes in mortality at any age.

Need to remove one of these:
- Due to data supression when death counts were small, we did not consider more specific causes of death such as homocide and drug overdose.

@@

###Results

@@

**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men (panel A) and women (panel B), in years, for each state. The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines. These lines are overlaid with the meta-estimated national trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S3-S14**).

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Figure 1A**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

@@

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**). Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**), with a peak contribution of `r can.male.max` in the early 1990s. For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a generally stable national pattern in the contribution of cancer to the life expectancy over time of approximately `r can.female.average` years (**Figure 2B**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

@@

Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**). Recently, there was a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**). As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men (approximately `r ncd.male.average` years), and exhibited a gradual decrease over time among women from `r ncd.female.1969` to `r ncd.female.2013` years (**Figures 3A and 3B**). As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress. Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap.

@@

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**). Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore decreased over time from `r inj.male.1969` to `r inj.male.2013` years among men. Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

```{r communicable_dx_stats, echo=F}
com.male.1969 <- cod_decomp_results %>% filter(year == 1969, sex == "Male" & COD == "Communicable" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

com.female.1969 <- cod_decomp_results %>% 
  filter(year == 1969, sex == "Female" & COD == "Communicable" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

com.male.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Male" & COD == "Cancers" & state == "Alabama") %>% 
  ungroup() %>%
  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))

com.female.mid.max <- cod_decomp_results %>% filter(year > 1980, sex == "Female" & COD == "Cancers" & state == "Alabama") %>% 
  ungroup() %>%
  select(fixed.from.RE, year) %>% summarize(max = round(max(fixed.from.RE), 1))

Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**). This spike was greater among men, especially blacks, and among states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the early 1990s on `r com.male.mid.max` years in men and `r com.female.mid.max` years in women, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

@@

```{r, echo = F}

aoc.male.endpoints <- cod_decomp_results %>% 
  filter(year %in% c(1969, 2013), sex == "Male" & COD == "All other causes" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

aoc.female.endpoints <- cod_decomp_results %>% 
  filter(year %in% c(1969, 2013), sex == "Female" & COD == "All other causes" & state == "Alabama") %>% 
  ungroup() %>% select(fixed.from.RE) %>% round(1)

aoc.male.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Male", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

aoc.female.endpoints.infant <- age_cod_results_infants_AOC %>% filter(year %in% c(1969, 2013), sex =="Female", state == "Alabama") %>% ungroup() %>% select(fixed.from.RE) %>% round(1)

proportions.males <- aoc.male.endpoints.infant/aoc.male.endpoints
proportions.females <- aoc.female.endpoints.infant/aoc.female.endpoints
```

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time (from `r sprintf("%.1f", aoc.male.endpoints[1, 1])` to `r aoc.male.endpoints[2, 1]` years among males and `r aoc.female.endpoints[1, 1]` to `r aoc.female.endpoints[2, 1]` years among females), and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**). The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally. We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**). We found that `r proportions.males[1,1]*100`% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969 in males and `r round(proportions.females[1,1]*100, 0)`% in females. In 2013, this proportion reached `r proportions.males[2,1]*100`% in males and `r proportions.females[2,1]*100`% in females. Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.

@@

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^16^

@@
